I would like to thank my thesis advisor, Jennifer Martin, for her constant encouragement, honest feedback and patience. She has read over my work from the very beginning, patiently sifting through ramblings of the creative, generative process while also encouraging me to explore various lines of inquiry. This resulted in lots of reading for her, but with her help much of that work has been refined in this thesis. Finally, I would like to thank her for helping me make the explicit connection between Job, Genesis and the \emph{Brothers Karamazov} as well as sharing many valuable insights with me.\\

I would also like to thank my friend, Dennis Monokroussos, who also has been a great help in the process. He directed my attention towards the work of Eleonore Stump, which served as a springboard for many of the ideas throughout the text such as the idea that the ``problem of suffering'' is an \emph{interpersonal} problem as opposed to a purely logical problem. \\

\pagebreak

\centerline{DEDICATION}

This thesis is dedicated to my mother, who is a constant source of love and whose perseverance is a great inspiration, and my deceased father, who taught me to see the world with tender, sensitive joy. In pouring themselves into their children, their sacrifice is representative of the way in which ``active love'', in the form of daily parental sacrifice, can yield great fruit and for that sacrifice I am tremendously thankful. \\

\begin{center}
{\emph{Verily, verily, I say unto you, Except a corn of wheat fall into the ground and die, it abideth alone: but if it die, it bringeth forth much fruit.} \\ John 12:24}
\end{center}

